---
title: "Creating State Maps"
author: "Rachel Carroll, Mazama Science"
date: "October 23, 2020"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

The goal of this document is to introduce the `stateMap()` function in the 
**MazamaSpatialPlots** package. It demonstrates default usage and customizations
using the `stateMap()` function's arguments.

## Default Plots

The `stateMap()` function requires two types of data. First is a 
simple features data frame (SFDF) containing state level polygons. The SFDF must 
include the variable `stateCode` in it's `` slot. The default SFDF is 
"USCensusStates_02", which part of the package data. Higher or lower 
resolution US Census state SFDFs can be installed with:

```
library(MazamaSpatialUtils)
setSpatialDataDir('~/Data/Spatial_0.8') # default directory for spatial data
installSpatialData('USCensusStates_02')
```

The second dataset is a regular dataframe that contains the 
variable `stateCode` as well as a variable of interest. The variable of interest 
from this dataset is indicated using the `parameter` argument. This `parameter`
is used to determine the colors of states in the generated chloropleth map. 

The next three examples demonstrate obtaining, summarizing, and mapping 
state-level data. The first two use example dataframes from the package and the 
third uses `MazamaCoreUtils::html_getTable()` to easily pull data from a website.

### Using package dataframe `example_US_stateObesity`

In this example, the package-internal dataframe, `example_US_stateObesity`, 
is used directly:

```{r plot_obesity, warning = FALSE, message = FALSE }
library(MazamaSpatialPlots)

stateMap(
  data = example_US_stateObesity,
  parameter = 'obesityRate',
  state_SFDF = "USCensusStates_02", # the default value
  title = "Obesity Rate in U.S. States"
)
```

### Using package dataframe `example_US_countyCovid`

In this example, the county-level package dataframe, `example_US_countyCovid`, 
is aggregated by state and then used:

```{r plot_covid, warning = FALSE, message = FALSE }
library(dplyr)

example_US_stateCovid <- 
  example_US_countyCovid %>%
  dplyr::group_by(stateCode) %>%
  dplyr::summarise(stateCases = sum(cases), stateDeaths = sum(deaths)) %>%
  dplyr::mutate(stateDeathRate = 100*stateDeaths/stateCases ) %>%
  dplyr::select(c("stateCode", "stateCases", "stateDeaths", "stateDeathRate"))

stateMap(
  data = example_US_stateCovid,
  parameter = 'stateDeathRate', 
  title =  "COVID Fatality Rates (%) on June 12, 2020"
)
```

### Using data found online

State-level data of interest can be found online and easily scraped using 
`MazamaCoreUtils::html_getTable()` to parse all table elements from a website. 
Using `html_getTable()` in conjunction with `stateMap()` makes it very easy to 
extract and visualize data from the internet.

```{r html_getTable_and_stateMap, warning = FALSE, message = FALSE }

URL <- "https://www.patriotsoftware.com/blog/accounting/average-cost-living-by-state/"
livingCostData <- MazamaCoreUtils::html_getTable(URL, header = TRUE)

livingCostData <- 
  livingCostData %>%
  dplyr::mutate(
    stateCode = MazamaSpatialUtils::US_stateNameToCode(.data$"State"),
    avgAnnualWage = as.numeric(gsub('[$,]', '', .data$"Annual Mean Wage (All Occupations)")),
    avgMonthlyRent = as.numeric(gsub('[$,]', '', .data$"Average Monthly Rent")),
    rentWagePercent = 100*12*avgMonthlyRent/avgAnnualWage,
    .keep = "none"
  ) 

stateMap(
  data = livingCostData,
  parameter = 'rentWagePercent',
  title = "Average Percentage of Income Spent on Rent"
)
```

## Customizing with Function Parameters

In the above examples, the `stateMap()` inputs `data`, `parameter`, 
`state_SFDF`, and `title` are used to create maps. This section will demonstrate 
how to use the other input parameters to customize your map. 

### Using palette, breaks, and stateBorderColor

The `palette`, `breaks`, and `stateBorderColor` parameters dictate the coloring 
of your map. Colors are defined with `palette` and the distribution of 
color across the map is defined with `breaks`. As expected, `stateBorderColor` 
defines the state border color. 

To make the most of these parameters, see the following references for R colors 
and palettes:

  * <http://sape.inf.usi.ch/quick-reference/ggplot2/colour>
  
  * <https://www.r-graph-gallery.com/38-rcolorbrewers-palettes.html>

In this example, `breaks` is used to create a coarser coloring scheme and 
`palette` is used to customize the exact color for each obesity rate level.
The vector of `breaks` will be one longer than the vector of `colors`.

```{r plotColors, warning = FALSE, message = FALSE }

stateMap(
  data = example_US_stateObesity,
  parameter = 'obesityRate',
  palette = c("lightblue", "orange", "red"),
  breaks = c(20, 27, 34, 40),
  stateBorderColor = "black",
  title = "Obesity Rate in U.S. States"
)

```

In this example, `breaks` is used to create a more detailed coloring scheme and 
the **RColorBrewer** blue to purple color palette name is chosen.

```{r plotStates, warning = FALSE, message = FALSE }

stateMap(
  data = example_US_stateObesity,
  parameter = 'obesityRate',
  palette = 'brewer.bu_pu',
  breaks = seq(20, 38, 3),
  stateBorderColor = 'white',
  title = "Obesity Rate in U.S. States"
)

```

### Using conusOnly and stateCode

The `conusOnly` and `stateCode` parameters define which states will be included 
in the map. If `stateCode` is defined, then `conusOnly` will be ignored. If 
`stateCode` is not defined, then `conusOnly` specifies whether the map is limited
to the continental US. When `conusOnly = FALSE`, the continental U.S., Alaska, 
Hawaii, and U.S. Territories will be included. 

This example builds upon the previous example and includes `stateCode` 
specification to create a map of Western states. 

```{r use_stateCode, warning = FALSE, message = FALSE }

stateMap(
  data = example_US_stateObesity,
  parameter = 'obesityRate',
  palette = 'brewer.pu_bu_gn',
  breaks = seq(23, 31, 2),
  stateCode = c("CA", "NV", "OR", "WA", "ID"),
  stateBorderColor = 'black',
  title = "Obesity Rate in Western U.S."
)


```

The following example uses `conusOnly = FALSE`.

```{r use_conusOnly, warning = FALSE, message = FALSE }

stateMap(
  data = example_US_stateObesity,
  parameter = 'obesityRate',
  palette = 'brewer.yl_or_rd',
  breaks = seq(20, 38, 3),
  conusOnly = FALSE,
  stateBorderColor = 'black',
  title = "Obesity Rate in U.S.",
  legendPosition = tmap::tm_pos_in("right", "top")
)


```

## Conclusion

The `stateMap()` function allows us to create attractive maps with a minimum of effort. 
When used alongside `MazamaCoreUtils::html_getTable()`, U.S. state data can be 
procured and visualized in very few lines of code. The combination 
of these two functions provides a great deal of flexibility. The `html_getTable()` 
function opens up an endless source of data from the internet while `stateMap()` 
can create highly customized visualizations through direct inputs and by 
[harnessing the functionality of the **tmap** package](Customizing_State_Maps.html). 






