---
title: "Creating County Maps"
author: "Mazama Science"
date: "February 4, 2020"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

The goal of this document is to introduce the `countyMap()` function in the 
**MazamaSpatialPlots** package. It demonstrates default usage and customizations 
using the `countyMap()` function's arguments.

## Default Plots

The `countyMap()` function requires three types of data. The first two 
are simple features data frames (SFDF) containing state and county level polygons. 
The SFDF containing state level polygons must include the variable `stateCode` 
in its `` slot. Similarly, the SFDF containing county level polygons must include
the variable `countyFIPS` in its `` slot. The default state and county level 
SFDF's are `USCensusStates_02` and `USCensusCounties_02` respectively, and are
part of the package data. Higher or lower resolution US Census state and county
SFDFs can be installed with:

```
library(MazamaSpatialUtils)
setSpatialDataDir('~/Data/Spatial_0.8')    # default directory for spatial data
installSpatialData('USCensusStates')   # state level polygons
installSpatialData('USCensusCounties') # county level polygons
```

The third dataset is a regular dataframe that contains the 
variable `countyFIPS` as well as a variable of interest. The variable of interest 
from this dataset is indicated using the `parameter` argument. This `parameter`
is used to determine the colors of counties in the generated chloropleth map. 

The next two examples demonstrate obtaining, summarizing, and mapping 
county-level data. The first uses an example dataframe from the package and the 
second demonstrates using `read.csv()` to pull data from the web and filling
in missing `countyFIPS` using `MazamaSpatialUtils::US_countyCodes`.

### Using package dataframe `example_US_countyCovid`

In this example, the package-internal dataframe, `example_US_countyCovid`, 
is used directly:

```{r plot_obesity, warning = FALSE, message = FALSE }
library(MazamaSpatialPlots)

countyMap(
  data = example_US_countyCovid,
  parameter = 'cases',
  breaks = c(0,100,200,500,1000,2000,5000,10000,20000,50000,1e6),
  state_SFDF = "USCensusStates_02", # the default value
  county_SFDF = "USCensusCounties_02", # the default value
  title = "Covid cases by county -- June 01, 2020"
)
```

### Using data found online

County level data of interest can be found online and easily loaded using 
`read.csv()` or scraped using `MazamaCoreUtils::html_getTable()` to parse 
all table elements from a website. 
Using these loading/scraping functions in conjunction with `countyMap()` makes 
it very easy to extract and visualize data from the internet.

```{r readcsv_and_coutymap, warning = FALSE, message = FALSE }

# readme for data: https://healthinequality.org/dl/health_ineq_online_table_12_readme.pdf
URL <- "https://healthinequality.org/dl/health_ineq_online_table_12.csv"
characteristicsData <- read.csv(URL)

# Added required 'stateCode' and 'countyFIPS' variables
characteristicsData <- 
  characteristicsData %>%
  dplyr::mutate(
    stateCode = stateabbrv,
    countyFIPS = MazamaSpatialUtils::US_countyNameToFIPS(stateCode, county_name),
    pUninsured2010 = puninsured2010,
    .keep = "none"
  ) 

# Create map
countyMap(
  data = characteristicsData,
  parameter = 'pUninsured2010',
  title = "Percentage of population uninsured in 2010"
)
```

## Customizing with Function Parameters

In the above examples, the `countyMap()` inputs `data`, `parameter`, 
`state_SFDF`, `county_SFDF` and `title` are used to create maps. This section 
will demonstrate how to use the other input parameters to customize your map. 

### Using palette, breaks, and stateBorderColor

The `palette`, `breaks`, and `stateBorderColor` parameters dictate the coloring 
of your map. Colors are defined with `palette` and the distribution of 
color across the map is defined with `breaks`. As expected, `stateBorderColor` 
defines the state border color. 

To make the most of these parameters, see the following references for R colors 
and palettes:

  * <http://sape.inf.usi.ch/quick-reference/ggplot2/colour>
  
  * <https://www.r-graph-gallery.com/38-rcolorbrewers-palettes.html>

In this example, `breaks` is used to create a coarser coloring scheme and 
`palette` is used to customize the exact color for each obesity rate level.
The vector of `breaks` will be one longer than the vector of `colors`.

```{r plotColors, warning = FALSE, message = FALSE }

countyMap(
  data = example_US_countyCovid,
  parameter = 'cases',
  palette = c("lightblue", "orange", "red"),
  breaks = c(0,1000,10000,50000,1e6),
  countyBorderColor = "white",
  title = "Covid cases by county -- June 01, 2020"
)

```

In this example, `breaks` is used to create a more detailed coloring scheme and 
the **RColorBrewer** blue to purple color palette name is chosen.

```{r plotStates, warning = FALSE, message = FALSE }

countyMap(
  data = example_US_countyCovid,
  parameter = 'cases',
  palette = 'BuPu',
  breaks = c(0,100,200,500,1000,2000,5000,10000,20000,50000,1e6),
  countyBorderColor = "white",
  title = "Covid cases by county -- June 01, 2020"
)

```

### Using conusOnly and stateCode

The `conusOnly` and `stateCode` parameters define which states will be included 
in the map. If `stateCode` is defined, then `conusOnly` will be ignored. If 
`stateCode` is not defined, then `conusOnly` specifies whether the map is limited
to the continental US. When `conusOnly = FALSE`, the continental U.S., Alaska, 
Hawaii, and U.S. Territories will be included. 

This example builds upon the previous example and includes `stateCode` 
specification to create a map of Western states. 

```{r use_stateCode, warning = FALSE, message = FALSE }


countyMap(
  data = example_US_countyCovid,
  parameter = 'cases',
  palette = 'BuPu',
  breaks = c(0,100,200,500,1000,2000,5000,10000,20000,50000,1e6),
  countyBorderColor = "white",
  stateCode = c("CA", "NV", "OR", "WA", "ID"),
  title = "Covid cases by county -- June 01, 2020"
)

```


## Conclusion

The `countyMap()` function allows us to create attractive maps with a minimum of effort. 
When used alongside loading and scraping functions such as `read.csv` and 
`MazamaCoreUtils::html_getTable()`, U.S. county data can be procured and visualized 
in very few lines of code. Similarly to `stateMap()`, county level choropleth maps
can be customized by [harnessing the functionality of the **tmap** package](Customizing_State_Maps.html). 






