---
title: "Customizing State Maps"
author: "Rachel Carroll, Mazama Science"
date: "July 25, 2020"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

The goal of this document is to serve as a tutorial for creating customized maps of U.S. states and territories by using custom projections or the **tmap** package. 

## Using projections

In general, the default projections in `stateMap()` are appropriate for common 
state groupings. However, the projection parameter can be used to override the 
default projections as needed. 

Below is an unusual group of states with the default projection.

```{r defaultprojection, warning = FALSE, message = FALSE }
library(MazamaSpatialPlots)

northernStates <- c("AK", "WA", "ID", "MT", "ND", "MN", "WI", 
                    "MI", "OH", "PA", "NY", "VT", "NH", "ME")

stateMap(
  data = example_US_stateObesity,
  parameter = 'obesityRate',
  palette = 'brewer.yl_or_rd',
  breaks = seq(23, 33, 2),
  stateCode = northernStates,
  stateBorderColor = 'black',
  title = "Obesity Rate in Northern U.S. States"
)

```

The projection is oddly rotated in this case. To remedy this, a manual projection 
can be defined: 

```{r manualprojection, warning = FALSE, message = FALSE }

myproj <- "+proj=lcc +lat_1=32.5 +lat_2=71.4 +lat_0=51.9 +lon_0=-102.9 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
stateMap(
  data = example_US_stateObesity,
  parameter = 'obesityRate',
  palette = 'brewer.yl_or_rd',
  breaks = seq(23, 33, 2),
  stateCode = northernStates,
  projection = myproj,
  stateBorderColor = 'black',
  title = "Obesity Rate in Northern U.S. States"
)
```

## Customizing with the **tmap** Package

One benefit of using `stateMap()` is that the output is a _ggplot2_ object. 
Therefore, maps can be further customized using functionality from the **tmap**
and **ggplot2** packages. Many visual adjustments including legend and title 
locations, fonts, and background colors can be made by simply appending your 
`stateMap()` plot with arguments in `tmap::tm_layout()`. Furthermore,
`tmap::tm_style()` can be used to leverage built in styles and 
`tmap::tm_compass()` can be used to add a compass to your map. The three 
examples below demonstrate how to build customized visualizations using both 
`stateMap()` parameters and **tmap** functionality.

### Using tmap::tm_layout()

```{r tmapStyling1, warning = FALSE, warning = FALSE, message = FALSE }

stateMap(
  data = example_US_stateObesity, 
  parameter = "obesityRate",
  breaks = seq(20,38,3), #increasing color detail
  conusOnly = FALSE ,
  stateBorderColor = 'black',
  legendPosition = tmap::tm_pos_in('right', 'top')
) +
  tmap::tm_layout(
    frame = TRUE,
    bg.color = "grey85", 
    inner.margins  = .05,
  ) +
  tmap::tm_title_out(
    text = 'Obesity Rates in U.S. States and Territories',
    fontface = 2,
    fontfamily = "serif"
  )

```

### Using tmap style "classic" and a compass

```{r tmapStyling2, warning = FALSE, message = FALSE }

stateMap(
  data = example_US_stateObesity, 
  parameter = "obesityRate", 
  breaks = seq(20,40,4),
) +
  tmap::tm_style(
    'classic'
  ) +
  tmap::tm_layout(
    inner.margins  = .08
  ) +
  tmap::tm_title_out(
    text = 'Obesity Rate by State',
    size = 1.2
  )
  tmap::tm_compass() 
  

```

### Fancy  New England regional map

```{r tmapStyling3, warning = FALSE, message = FALSE }
  
stateMap(
  data = example_US_stateObesity,
  parameter = "obesityRate",
  stateCode = c('ME', 'NH', 'VT', 'MA', 'RI', 'CT'),
  stateBorderColor = 'black',
) +
  tmap::tm_layout(
    frame = TRUE,
    frame.double.line = TRUE,
    bg.color = "grey85",
    inner.margins  = .08
  ) +
  tmap::tm_title_out(
    text = 'Obesity Rates in New England',
    size = 1.2,
    fontface = 2,
    fontfamily = "serif"
  )

```

## Conclusion

The `stateMap()` function allows us to create attractive maps very efficiently 
and with a lot of flexibility. The function creates highly customized 
visualizations through direct inputs and by harnessing the functionality of the 
**tmap** package. 






