---
title: "Working with Air Quality Data"
author: "Rachel Carroll, Mazama Science"
date: "August 7, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

The objective of this article is to demonstrate how to make detailed maps from 
air quality monitoring data. The **PWFSLSmoke** package is used to extract and 
manipulate the air quality monitoring data from AIRNOW, AIRSIS and WRCC. This 
data contains particulate matter 2.5 (PM2.5) concentrations at the monitor 
locations over time. In this demonstration, we create maps that display monitor 
locations, air quality statistics at monitor locations, and average air quality 
in a given region. 

## Prepare Monitor Data

There are several functions in the **PWFSLSmoke** package that can be used to load air quality monitoring data:

  *  `monitor_load()` - Loads data from a given time range. For example, `monitor_load(20170601, 20171001)` brings in all readings between 06/01/17 and 10/01/17.
  *  `monitor_loadAnnual()` - Loads all data from a given year. For example `monitor_loadAnnual(2018)` pulls in all 2018 data.
  *  `monitor_loadDaily()` - Loads the past 45 days of data, up to midnight of the most recent full day.
  *  `monitor_loadDaily()` - Loads the past 10 days of data, up the current hour.
  
  
These functions load in the data as a _ws_monitor_ object, which contains 1) a dataframe with air quality monitor readings over time and 2) the monitor-level metadata with characteristics such as monitor locations. 

In the example below, data from the past 10 days is loaded and monitor-level 
calculations are performed. The data can be summarized in any way deemed relevant 
so long as we end up with one measurement per monitor. This allows us to color monitor 
location points based on the monitor's PM2.5 value reading. To accomplish this, 
the data can be summarized over time or data can be extracted at a specific 
point in time. In this case, the former method is used to calculate the maximum 
daily average for each monitor. 

```{r prepdata, warning = FALSE, message = FALSE}

library(MazamaSpatialPlots)
library(PWFSLSmoke)

# Load data
monitorLatestData <- PWFSLSmoke::monitor_loadLatest() # past 10 days of data, up to the current hour

# Prepare Data: Subset and summarize based on desired information
# calculating maximum daily average (one record per monitor)
dailyMax <-
  monitorLatestData %>%
  PWFSLSmoke::monitor_dailyStatistic(minHours = 2) %>% # daily average
  PWFSLSmoke::monitor_extractData() %>% 
  tidyr::pivot_longer(-datetime, names_to = "monitorID", values_to = "value") %>%
  dplyr::group_by(monitorID) %>% 
  dplyr::summarise(pm25 = max(value, na.rm = TRUE)) %>% # taking maximum daily average
  dplyr::left_join(monitorLatestData$meta, by = 'monitorID') %>%
  dplyr::select(
    c('monitorID', 'stateCode', 'latitude', 'longitude', 'pwfslDataIngestSource', 'pm25'))

# Clean pm25 column by setting -Inf values to NA
dailyMax$pm25 <- replace(dailyMax$pm25, dailyMax$pm25 == -Inf, NA)

```

Now the _ws_monitor_ object, `monitorLatestData` and the monitor-level dataframe, 
`dailyMax`, can be used to create air quality maps as demonstrated in the 
following sections.

## Plot Monitor Locations 

To simply plot monitor locations without colors, monitor-level calculations are 
not yet needed. Instead, the metadata of `monitorLatestData` can be used directly. 
We also introduce a U.S. state SpatialPolygonsDataFrame (SPDF) so the monitor 
locations will appear on a map of U.S. states.

For simplicity, we won't project the data at all.

```{r monitor_locs, warning = FALSE, message = FALSE}

# Get state codes and only include continental US states
stateCodeList <- unique(monitorLatestData$meta$stateCode) 
stateCodeList <- subset(
  stateCodeList, stateCodeList %in% MazamaSpatialUtils::CONUS
  )
# Create subsetted U.S. state SPDF
stateSPDF <- subset(USCensusStates_02, stateCode %in% stateCodeList)

# Plot U.S. state SPDF
plot(stateSPDF)  

# Overlay monitor location points
# NOTE: can use either dailyMax or monitorLatestData$meta since both have metadata 
points(monitorLatestData$meta$longitude, monitorLatestData$meta$latitude)
points(monitorLatestData$meta$longitude, monitorLatestData$meta$latitude)

```

Another method to map monitor locations is to use the `monitor_map()` function 
from **PWFSLSmoke**. In fact, this function can also color the monitor location 
points (by default, the maximum measurement per monitor is used to color the 
points). This function is very useful when taking a quick look across the 
continental U.S. However, the method above allows for some more flexibility in 
subsetting states and calculating desired monitor-level statistics.     

``` {r monitor_locs2, warning = FALSE, message = FALSE}
# method 2 using monitor_map() from PWFSLSmoke
PWFSLSmoke::monitor_map(
  monitorLatestData, 
  countyCol = NA,
  colors = 'grey50',
  stateCol = "black",
  pch = 3,
  stateLwd = 3
)
```                       

## Plot Monitor Values

To build upon the examples above, this section shows how to plot points on a map 
with added visual customizations.  To create descriptive visualizations of this 
nature, it is a good idea to transform the locations data into a 
_SpatialPointsDataFrame_. This is because SPDFs are compatible with the **tmap** 
package, which allows us to create attractive maps with high flexibility.

Below, monitor locations are again plotted on a plain continental U.S. map. In this example, however, **tmap** is used so the overall layout, color schemes, legend break points, and point shapes are easily specified.  Notice that AIRNOW monitors are displayed as circles and AIRSIS and WRCC monitors are triangles. This is done by simply defining the shape parameter in `tm_symbols()` with the data source column. This cannot be done using `PWFSLSmoke::monitor_map()` and would require longer messier code to create directly using `plot()` and `points()`.

```{r monitor_vals, warning = FALSE, message = FALSE}

# transform monitorLatestData into a SpatialPointsDataFrame 
# TODO: make this a function called monitor_toSPDF ?????
geojson_file <- tempfile(fileext = ".geojson")
current_geojson <- monitor_writeCurrentStatusGeoJSON(monitorLatestData, geojson_file)
current_list <- jsonlite::fromJSON(current_geojson)
monitorSPDF <- rgdal::readOGR(dsn = geojson_file, verbose = FALSE)

# bring in pm25 field from dailyMax 
monitorSPDF@data <- monitorSPDF@data %>%
  dplyr::left_join(dailyMax[, c("monitorID", "pm25")], by = "monitorID") %>%
  dplyr::mutate(  
    plotDataSource = ifelse(
      pwfslDataIngestSource == "AIRNOW", 
      "AIRNOW", 
      "AIRSIS or WRCC")
    ) %>%
  dplyr::select(
    c("monitorID", "stateCode", "longitude", "latitude", "plotDataSource", "pm25")
    )

#createMap
tmap::tm_shape(stateSPDF, 
               bbox = bbox(monitorSPDF) #use monitor bbox 
               ) +
  tmap::tm_polygons(
    col = "grey85",
    border.col = "black") +
  tmap::tm_shape(monitorSPDF) +  # plot colored monitor location points 
  tmap::tm_symbols(              
    col = "pm25",
    palette = "Reds",
    breaks = c(seq(0, 50, 5), 1000), # NOTE: all values are between 0 and 50 except 2 (outliers?)
    shape = "plotDataSource",        # shape of points will depend on the monitor source
    shapes = c(16, 17),              # specify circles and triangle shapes
    scale = .4) +
  tmap::tm_layout(
    main.title = "Latest Average Daily Max Monitor Reading",
    main.title.position = c("center", "top"),
    bg.color = "deepskyblue4",
    attr.color = "white",
    legend.position = c("left", "center")
  )

```

## Plot Polygon Summarized Values

In this section `MazamaSpatialUtils::summarizeByPolygon()` is introduced. The 
purpose of this function is to summarize spatial data by regions defined by the 
polygons in a given SPDF. For example, using a U.S. state level SPDF would 
summarize the data by state while a U.S. county SPDF would summarize by county. 
The input spatial data must contain latitude, longitude, and a value to be summarized. 

Below, `MazamaSpatialUtils::summarizeByPolygon()` is used with a U.S. state SPDF 
and the dataframe `dailyMax` (created in the "Prepare Monitor Data" section above) 
to find average PM2.5 readings by state. These state-level values are 
incorporated into the state SPDF's data `@slot` so the map can be colored by the 
summarized values.  

```{r plot_summarized, warning = FALSE, message = FALSE}
# replacing NA with 0. 
# NOTE: should add na.rm argument to summarizeByPolygon() so we can 
# NOTE: use dailyMax$pm25 directly in the function. This step will bring down 
# NOTE: the average value which is NOT GOOD 
stateSPDF <- subset(USCensusStates_02, stateCode %in% stateCodeList)

newValueVector <- replace(dailyMax$pm25, is.na(dailyMax$pm25), 0)

# Summarize by State
stateSummarizedDF <- MazamaSpatialUtils::summarizeByPolygon(
  longitude = dailyMax$longitude,
  latitude = dailyMax$latitude,
  value = newValueVector, 
  SPDF = stateSPDF,
  useBuffering = TRUE,
  FUN = mean
)

# add summarized values to stateSPDF@data
stateSPDF@data <- stateSPDF@data %>%
  dplyr::left_join(stateSummarizedDF, by = "polygonID") %>%
  dplyr::mutate(avgPM25 = summaryValue)

#createMap
tmap::tm_shape(stateSPDF) +
  tmap::tm_polygons(
    col = "avgPM25",
    breaks = c(seq(5, 13, 2), seq(15, 70, 5)),
    border.col = "black") +
  tmap::tm_layout(
    main.title = "Latest Average Daily Max Monitor Reading",
    main.title.position = c("center", "top"),
    bg.color = "deepskyblue4",
    attr.color = "white",
    legend.position = c("right", "bottom")
  )

```
 
## Sprinkle Points on Summarized Value Maps

In this section, we use both the state-level and monitor-level SPDFs to create a 
map with states colored by average air quality and monitor location points 
colored and sized by the maximum daily average PM2.5 concentrations

In the example below `MazamaSpatialPlots::stateMap()` is used for the state-level 
map and colors. Various **tmap** functions are added to overlay the monitor 
location points. 

```{r plot_summarized_points, warning = FALSE, message = FALSE}

#subset SPDFs
selectedStateCodeList <- c('ME', 'NH', 'VT', 'MA', 'RI', 'CT')
selectedStateSPDF <- subset(stateSPDF, stateCode %in% selectedStateCodeList)
selectedMonitorSPDF <- subset(monitorSPDF, stateCode %in% selectedStateCodeList)

# Now use different size markers since all are AIRNOW in New England
stateMap(
  data = selectedStateSPDF@data,
  parameter = "avgPM25", # color states by average PM Value
  stateCode = selectedStateCodeList,
  stateBorderColor = 'white', 
  breaks = seq(1, 20, 2),
) +
  tmap::tm_shape(selectedMonitorSPDF) +  # plot points of each monitor location with color 
  tmap::tm_bubbles(          # and size based on PM value reading of the monitor
    size = "pm25",
    col = "pm25", 
    border.col = 'black',
    breaks = seq(1, 20, 2),
    scale = .9,
    legend.size.show = F,
    legend.col.show = F
    ) +
  tmap::tm_layout(
    main.title = 'Air Quality in New England',
    main.title.size = 1.1,
    fontfamily = "serif",
    bg.color = "lightblue3",
  )
  
```

This next example only uses **tmap** functions. The `tm_symbols()` function is 
used to plot the monitor locations so the point shapes can be specified. 

```{r plot_summarized_points2, warning = FALSE, message = FALSE}

#subset SPDFs
selectedStateCodeList <- c("CA", "OR", "WA", "NV", "ID", "UT", "AZ")
selectedStateSPDF <- subset(stateSPDF, stateCode %in% selectedStateCodeList)
selectedMonitorSPDF <- subset(monitorSPDF, stateCode %in% selectedStateCodeList)

# can also do it with just tmap
tmap::tm_shape(selectedStateSPDF) +
  tmap::tm_polygons(
    col = "avgPM25", 
    palette = "YlOrBr",
    border.col = "black",
    breaks = seq(0, 50, 5),
    legend.show = FALSE) +
  tmap::tm_shape(selectedMonitorSPDF) +  # plot points of each monitor location with color 
  tmap::tm_symbols(          # and size based on PM value reading of the monitor
    col = "pm25",
    palette = "YlOrBr",
    breaks = seq(0, 50, 5),
    shape = "plotDataSource",
    shapes = c(21, 24),
    scale = .6
  ) +
  tmap::tm_layout(
    main.title = "Air Quality in Western U.S. States",
    main.title.position = c("left", "top"),
    main.title.size = 1.1,
    fontfamily = "serif",
    legend.outside = T,
    bg.color = "lightblue3", 
    inner.margins = .1,
  )
```

## Conclusion

Using a few key functions from **MazamaSpatialUtls**, **MazamaSpatialPlots**
and **PWFSLSmoke** makes it easy to create detailed and descriptive maps 
containing both regional averages as well as spatial data point measurements. 
The examples in this article demonstrate several ways to leverage these functions
as well as customization capabilities from the **tmap** package.



